Global Life Expectancy: Unraveling Health and Economic Determinants (2000-2015)

### 2024-11-28 paper/paper.qmd
data cleaning

---
title: "Global Life Expectancy: Unraveling Health and Economic Determinants (2000-2015)"
subtitle: "Analyzing Critical Factors Shaping Mortality and Well-Being Across 193 Nations"
author: 
  - Yanfei Huang
thanks: "Code and data are available at: [https://github.com/wendyhuan/lifeexpectancy](https://github.com/wendyhuan/lifeexpectancy)."
date: 26 November 2024
date-format: long
abstract: "This paper analyze life expectancy and its determinants through 193 countries and make the prediction of people from both developed and developing country. Multiple Linear regression is used to deploying the health and socio-economic factors. Predictions of life expectancy of people from developed or developing country is made according to the essential predictors. The finding indicates that the developed countries tend to live longer life as assumed. And we predict that the average age of persons in developed country is 80, while those in underdeveloped countries are 75. The result of this paper will help in suggesting a country which area should be given importance in order to efficiently improve the life expectancy of its population."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

# Install Packages if not downloaded:
required_packages <- c(
  "tidyverse", "sf", "dplyr", "here", "arrow", "ggplot2", "janitor", "purrr", "knitr", "kableExtra", "ggcorrplot", "modelsummary"
)
for (p in required_packages) {
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}

@@

# Introduction
Life expectancy is a critical measure of a population's overall health and well-being, shaped by various factors such as personal expenditure on health, government healthcare access, economic stability, and social structures. The index of life expectancy is generally served as a benchmark for development, indicating the effectiveness of interventions in reducing mortality and improving well-being. Higher life expectancy is believed to linked to better living standards, access to education, and equitable healthcare. Conversely, low life expectancy often signals systemic challenges such as poverty, disease burden, and inadequate healthcare access [@socialdeterminantsofhealth].

@@

The data used in this analysis originates from The World Health Organization’s (WHO) and Global Health Observatory (GHO), as compiled and merged by Kumar Rajarshi on Kaggle. [@lifeexpectancykaggle]. The analysis uses the statistical programming language R [@citeR] and several libraries, including `tidyverse` [@tidyverse], `janitor` [@janitor], `knitr` [@knitr], `dplyr` [@dplyr], `arrow` [@arrow], `purrr` [@purrr], `sf` [@sf], and `here` [@here] for data manipulation. `ggplot2` [@ggplot], `ggcorrplot` [@ggcorrplot] and `kableExtra` [@kableExtra] for visualization. The dataset covers various predictors conducted across multiple countries, capturing the support for a country to determine the predicting factor which is contributing to lower value of life expectancy.

@@

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

@@

@@

# Appendix {-}

>>>>

---
title: "Global Life Expectancy: Unraveling Health and Economic Determinants (2000-2015)"
subtitle: "Analyzing Critical Factors Shaping Mortality and Well-Being Across 193 Nations"
author: 
  - Yanfei Huang
thanks: "Code and data are available at: [https://github.com/wendyhuan/lifeexpectancy](https://github.com/wendyhuan/lifeexpectancy)."
date: 30 November 2024
date-format: long
abstract: "This paper analyze life expectancy and its determinants through 193 countries and make the prediction of people from both developed and developing country. Multiple Linear regression is used to deploying the health and socio-economic factors. Predictions of life expectancy of people from developed or developing country is made according to the essential predictors. The finding indicates that the developed countries tend to live longer life as assumed. And we predict that the average age of persons in developed country is 80, while those in underdeveloped countries are 75. The result of this paper will help in suggesting a country which area should be given importance in order to efficiently improve the life expectancy of its population."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

# Install Packages if not downloaded:
required_packages <- c(
  "tidyverse", "sf", "dplyr", "here", "arrow", "ggplot2", "janitor", "purrr", "knitr", "kableExtra", "ggcorrplot", "modelsummary", "countrycode"
)
for (p in required_packages) {
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}

@@

# Introduction

Life expectancy is a critical measure of a population's overall health and well-being, shaped by various factors such as personal expenditure on health, government healthcare access, economic stability, and social structures. The index of life expectancy is generally served as a benchmark for development, indicating the effectiveness of interventions in reducing mortality and improving well-being. Higher life expectancy is believed to linked to better living standards, access to education, and equitable healthcare. Conversely, low life expectancy often signals systemic challenges such as poverty, disease burden, and inadequate healthcare access [@socialdeterminantsofhealth].

@@

The data used in this analysis originates from The World Health Organization’s (WHO) and Global Health Observatory (GHO) [@lifeexpectancy]. This data-set related to life expectancy, health factors for 193 countries has been collected from WHO data repository website and its corresponding economic data was collected from United Nation website. Among all categories of health-related factors only those socio-economic factors on the national level were chosen for global scale analysis.

This analysis uses the statistical programming language R [@citeR] and several libraries, including `tidyverse` [@tidyverse], `janitor` [@janitor], `knitr` [@knitr], `dplyr` [@dplyr], `arrow` [@arrow], `purrr` [@purrr], `sf` [@sf], and `here` [@here] for data manipulation. `ggplot2` [@ggplot], `ggcorrplot` [@ggcorrplot] and `kableExtra` [@kableExtra] for visualization. The dataset covers various predictors conducted across multiple countries, capturing the support for a country to determine the predicting factor which is contributing to lower value of life expectancy.

@@

The measurement process refers to how real-world factors—such as the status of a country, average personal health care investment, average year of education,  geographic location of a country and social determinants (country's healthcare investment, human development index of income) — are translated into numerical entries representing life expectancy in a dataset. Each entry captures the average life expectancy of individuals in a specific country during a given year.

Life Expectancy (`LifeExpectancy`): This variable represents the average number of years a person is expected to live, assuming current mortality conditions persist. It is derived using data from national health records, the World Health Organization (WHO) and Global Health Observatory (GHO). The values are calculated and represented in age between 36.3 to 89.

Percentage Expenditure on Health (`PercentageExpenditure`): This variable reflects the expenditure allocated to health-related expenditures as a percentage of GDP per capita. The data is typically collected from government reports, economic surveys, and global health databases. It indicates the level of personal financial investment in the health sector and is expressed as a percentage.

Schooling (`Schooling`): This variable captures the average number of years of education received by individuals in a country. Data is obtained from educational surveys and censuses conducted by international and national organizations. The values are recorded in years and used to assess the link between education and health outcomes like life expectancy. The data range from 0 to 20.7

Status (`Status`): This is a categorical variable that classifies countries as "Developed" or "Developing," based on various socio-economic and health indicators accepted globally. It provides context for comparing life expectancy across different stages of national development.

Continent(`Continent`): This is a categorical variable that classifies the geographic location of the countries. It is detected by the name of the country using rpackage `countrycode`. The data is stored as the name of the continent(‘Africa’, ‘Oceania’, ‘Asia’, ‘Europe’, ‘North America’, ‘South America’)

Total expenditure (`TotalExpenditure`): This is a numerical variable that shows the general government expenditure on health as a percentage of total government expenditure. The number drop of the percentage and shows as the number between 0 to 100, where higher values indicate a greater share of government spending allocated to health.

Income composition of resource (`IncomeComposition`): This is a numerical variable representing the income component of the Human Development Index (HDI). It reflects the contribution of income to human development, with values ranging from 0 to 1, where higher values signify a greater income contribution to overall development.

@@

The raw life expectancy data underwent a several cleaning steps to ensure it was accurate, consistent and ready for analysis. We first select and rename key variables from raw data to focus on relevant information. The key variables of interest in our analysis include Country, Year, Status of the country, Life expectancy, Percentage expenditure, Total expenditure, Income composition of resources and the year of education. Additionally, we develop a new predictor named "Continent" to monitor each nation's geographic data. To make the subsequent analysis easier, we then convert variables to the proper data types and eliminate the rows that have missing data values. To keep things neat, we organize the decimal for every piece of numerical data. The cleaned dataset was then saved as a Parquet file for efficient storage and further analysis.

More information on the data cleaning process can be found in @sec-appx.

@@

# Appendix {#sec-appendix}

## Data Cleaning Notes

We began by importing the raw dataset using the read_csv function from the tidyverse package. To focus our analysis on more relevant variables, we selected specific columns, such as percentage of expenditure of the country, total health expenditure by government .etc., omitting any unnecessary columns.

We filter out the rows containing NA values in any of the selected columns for reducing the noise and simpler further analysis..

We then renaming columns for clarity. For example, we changed ‘Income.composition.of.resources’ into ‘IncomeComposition’ , making it easier for anyone working with the data to read and understand what each variable represents.

Each column is rounded using the round function, specifying the desired number of decimal places for each. Columns not mentioned in mutate remain unchanged. This ensures a flexible and precise cleaning process tailored to further compairation and graphing.

We also modified the country name of ‘Republic of Korea’ under Country variable to ‘South Korea’ to avoid misunderstanding of North Korea.

We also created a new variable called Continent, which indicates which continent does the country comes from in order to provide a geographical context to the analysis. Life expectancy may be higher in developed regions like Europe or North America compared to regions like Sub-Saharan Africa due to differences in healthcare, living standards, and economic development.

We also wrapped the cleaning process in a tryCatch block in order to mitigate any errors that arose throughout the cleaning process.

After completing the cleaning, we saved the final dataset in both Parquet and CSV formats for later analysis.
